---
title: "RNA-seq Worksheet: Differential Expression Analysis with the airway Dataset"
author: ""
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 2
    number_sections: true
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Overview

In this worksheet, you will analyze a small RNA-seq dataset using **DESeq2**. The dataset comes from the Bioconductor package **airway** and contains RNA-seq counts from **human airway smooth muscle cell lines** under two conditions:

- **untrt** = untreated
- **trt** = treated with dexamethasone

Your goal is to:

1. load the dataset,
2. check samples and metadata,
3. set up and run a DESeq2 analysis,
4. create standard RNA-seq plots,
5. identify top differentially expressed genes,
6. write a brief interpretation of the results.

# Learning objectives

By the end of this activity, you should be able to:

- import RNA-seq counts and metadata from a Bioconductor dataset,
- verify that samples and metadata match,
- create a `DESeqDataSet`,
- run differential expression analysis with DESeq2,
- interpret PCA, MA, and volcano plots,
- summarize top genes and biological patterns.

# Before you begin

Install packages if needed.

```{r install-packages, eval=FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE)) {
  install.packages("BiocManager")
}

BiocManager::install(c("airway", "DESeq2", "SummarizedExperiment"), ask = FALSE, update = FALSE)
install.packages(c("ggplot2", "pheatmap"))
```

Load required libraries.

```{r load-libraries}
library(airway)
library(DESeq2)
library(SummarizedExperiment)
library(ggplot2)
library(pheatmap)
```

# Part 1. Load the dataset

Load the `airway` dataset.

```{r load-airway}
data(airway)
airway
```

Extract the count matrix and metadata.

```{r extract-data}
counts <- assay(airway)
meta <- as.data.frame(colData(airway))

head(counts[, 1:4])
meta
dim(counts)
```

## Questions

1. How many genes are in the dataset?  
   63677 genes in this dataset

2. How many samples are in the dataset?  
   There are 8 samples 

3. What are the two treatment groups?  
   Treated and un-treated

# Part 2. Sample check and metadata check

Check the dimensions of the count matrix.

```{r dimensions}
dim(counts)
```

Check sample names in the count matrix.

```{r sample-names}
colnames(counts)
```

Check sample names in the metadata.

```{r metadata-names}
rownames(meta)
```

Confirm that the sample names match.

```{r match-check}
all(colnames(counts) == rownames(meta))
```

Inspect important metadata columns.

```{r metadata-summary}
meta[, c("cell", "dex")]
table(meta$dex)
table(meta$cell)
```

## Questions

4. Do the sample names in the count matrix and metadata match?  
   Yes the row name of the metadata matches the column name of the count matrix 

5. How many samples are treated and how many are untreated?  
   4 samples treated and 4 samples untreated

6. How many unique cell lines are included?  
   four unique cell lines 

# Part 3. Quick quality checks

## 3.1 Library sizes

Calculate the total counts per sample.

```{r library-sizes}
lib_sizes <- colSums(counts)
lib_sizes
```

Plot library sizes.

```{r library-barplot}
barplot(lib_sizes,
        las = 2,
        main = "Library sizes",
        ylab = "Total counts")
```

## 3.2 Gene count distribution

Check total counts across genes.

```{r gene-sums}
summary(rowSums(counts))
```

Plot the distribution of total counts per gene.

```{r gene-sums-hist}
hist(rowSums(counts),
     breaks = 100,
     main = "Total counts per gene",
     xlab = "Row sums")
```

## Questions

7. Do the library sizes look similar across samples, or are there large differences?  
   Library size look similar across samples

8. Why might very low-count genes be filtered before differential expression analysis?  
   DESeq assumes some level of expression from genes, even if it's low, otherwise it
   coule be becuase sequencer didn't detect rather than low expression

# Part 4. DESeq2 setup

Create a DESeq2 dataset. Because treatment is measured within cell lines, we will model:

```r
~ cell + dex
```

This means we test the treatment effect while adjusting for cell-line differences.

```{r dds-setup}
dds <- DESeqDataSet(airway, design = ~ cell + dex)
dds
```

Set untreated as the reference group.

```{r relevel}
dds$dex <- relevel(dds$dex, ref = "untrt")
levels(dds$dex)
```

Optional filtering step: keep genes with at least 10 counts in at least 2 samples.

```{r filtering}
keep <- rowSums(counts(dds) >= 10) >= 2
dds <- dds[keep, ]
dds
```

## Questions

9. Why do we include `cell` in the design formula?  
   because each cell type is repeated twice, trt and untrt, and must be grouped accordingly rather than by sample

10. What does it mean to set `untrt` as the reference level?  
   untrt read count is the baseline, used to compare the significance of differential gene expression

# Part 5. Run DESeq2

Run the differential expression pipeline.

```{r run-deseq}
dds <- DESeq(dds)
```

Extract results for treated vs untreated.

```{r get-results}
res <- results(dds, contrast = c("dex", "trt", "untrt"))
summary(res)
```

Convert results to a data frame and sort by adjusted p-value.

```{r sort-results}
res_df <- as.data.frame(res)
res_df <- res_df[order(res_df$padj), ]
head(res_df)
```

Count significant genes.

```{r significant-genes}
sig <- subset(res_df, padj < 0.05)
nrow(sig)
```

## Questions

11. How many genes are significant at adjusted p-value < 0.05?  
   4100

12. Why do we use adjusted p-values instead of raw p-values?  
   to account for false discovery rate (FDR)

# Part 6. PCA plot

Use a variance-stabilizing transformation for PCA.

```{r vst}
vsd <- vst(dds, blind = FALSE)
```

Create a PCA plot.

```{r pca-plot}
plotPCA(vsd, intgroup = c("dex", "cell"))
```

Optional custom PCA plot.

```{r custom-pca}
pcaData <- plotPCA(vsd, intgroup = c("dex", "cell"), returnData = TRUE)
percentVar <- round(100 * attr(pcaData, "percentVar"))

ggplot(pcaData, aes(PC1, PC2, color = dex, shape = cell)) +
  geom_point(size = 4) +
  xlab(paste0("PC1: ", percentVar[1], "% variance")) +
  ylab(paste0("PC2: ", percentVar[2], "% variance")) +
  ggtitle("PCA of airway samples")
```

## Questions

13. Do samples appear to separate by treatment, cell line, or both?  
   samples appear to separate by treatment more significantly than cell line 

14. Are there any possible outliers?  
   No

# Part 7. MA plot

Create an MA plot.

```{r ma-plot}
plotMA(res, ylim = c(-5, 5))
```

## Question

15. What does the MA plot show about gene expression changes between treated and untreated samples?  
   there are genes that are both up and down regulated significantly

# Part 8. Volcano plot

Create a volcano plot using DESeq2 results.

```{r volcano-data}
volcano_df <- as.data.frame(res)
volcano_df$gene <- rownames(volcano_df)
volcano_df <- volcano_df[!is.na(volcano_df$padj), ]

volcano_df$significance <- "Not significant"
volcano_df$significance[volcano_df$padj < 0.05 & volcano_df$log2FoldChange > 1] <- "Up"
volcano_df$significance[volcano_df$padj < 0.05 & volcano_df$log2FoldChange < -1] <- "Down"

print(length(volcano_df$significance[volcano_df$padj < 0.05 & volcano_df$log2FoldChange > 1]))
print(length(volcano_df$significance[volcano_df$padj < 0.05 & volcano_df$log2FoldChange < -1]))
```

```{r volcano-plot}
ggplot(volcano_df, aes(x = log2FoldChange, y = -log10(padj), color = significance)) +
  geom_point(alpha = 0.7) +
  theme_minimal() +
  labs(
    title = "Volcano plot: treated vs untreated",
    x = "log2 fold change",
    y = "-log10 adjusted p-value"
  )
```

Optional: label the top 10 genes.

```{r volcano-labels}
top_lab <- head(volcano_df[order(volcano_df$padj), ], 10)

ggplot(volcano_df, aes(x = log2FoldChange, y = -log10(padj), color = significance)) +
  geom_point(alpha = 0.7) +
  geom_text(data = top_lab, aes(label = gene), vjust = -0.5, size = 3, show.legend = FALSE) +
  theme_minimal() +
  labs(
    title = "Volcano plot with top genes labeled",
    x = "log2 fold change",
    y = "-log10 adjusted p-value"
  )
```

## Questions

16. Approximately how many strongly upregulated and downregulated genes do you observe?  
   531 upregulated and 497 downregulated genes

17. What information does the volcano plot show that complements the MA plot?  
   idfk

# Part 9. Top genes and heatmap

Add gene names as a column and inspect the top significant genes.

```{r top-genes}
sig$gene <- rownames(sig)
sig <- sig[, c("gene", colnames(sig)[!colnames(sig) %in% "gene"])]
head(sig, 10)
```

Save results if desired.

```{r save-results, eval=FALSE}
write.csv(as.data.frame(res_df), "airway_deseq2_results.csv", row.names = TRUE)
write.csv(sig, "airway_significant_genes.csv", row.names = FALSE)
```

Create a heatmap of the top 20 significant genes.

```{r heatmap}
top_genes <- rownames(subset(res_df, padj < 0.05))[1:20]
mat <- assay(vsd)[top_genes, ]
mat <- mat - rowMeans(mat)

print(top_genes)

pheatmap(mat,
         annotation_col = meta[, c("dex", "cell"), drop = FALSE],
         main = "Top 20 DE genes")
```

## Questions

18. List the top 5 significant genes.  
   "ENSG00000152583" "ENSG00000165995" "ENSG00000120129" "ENSG00000101347" "ENSG00000189221"

19. Does the heatmap show clear clustering by treatment, cell line, or both?  
   Heatmap shows clear clustering by treatment not cell line

# Part 10. Short interpretation

Write a short paragraph summarizing your findings.

Use the prompts below:

- What comparison did you perform?
- Why did you include `cell` in the model?
- Did the PCA suggest separation by treatment, cell line, or both?
- Were many genes significantly differentially expressed?
- What is one biological conclusion you can make from this analysis?

## Your interpretation

**Write 4 to 6 sentences here:**



# Submission checklist

Make sure your submission includes:

- completed code chunks,
- answers to the questions,
- PCA plot,
- MA plot,
- volcano plot,
- heatmap,
- short interpretation paragraph.

# Optional challenge

Try one of the following:

1. Count how many genes are upregulated and downregulated at `padj < 0.05` and `|log2FC| > 1`.
2. Make a table of the top 20 genes.
3. Compare results before and after filtering low-count genes.
4. Look up one top gene and describe its possible role in glucocorticoid response.
